Empirical Bayes analysis of RNA-seq data for detection of gene expression heterosis

### 2015-03-02 ebayes_rnaseq_heterosis.tex
added publisher
post Dan edits, but pre comment fixes

\newcommand{\blind}{1} % 0 = not blind, 1 = blind

\title{Empirical Bayes analysis for detection of gene heterosis in RNAseq data}

@@

\begin{abstract}
Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents. Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis is still unknown. In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to compare expression levels of thousands of genes between parental inbred lines and their hybrid offspring to search for evidence of gene expression heterosis in RNA sequencing (RNAseq) experiments. Standard statistical approaches for RNAseq experiments struggle with the composite null hypothesis in gene heterosis testing. To address these shortcomings, we develop a hierarchical model based on a negative-binomial data model and a computationally tractable empirical Bayes approach. We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment. This article has supplementary material online.
\end{abstract}

\begin{keywords}
Empirical Bayes; Gene expression; Heterosis; Hierarchical model; Negative binomial; RNAseq.
\end{keywords}

@@

% Need to udpate as this is directly from Tieming's paper
Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents. Taking maize as an example, the offspring from crossing the inbred lines B73 and Mo17 grow taller, mature faster, and produce greater yields than their parent lines \citep{hallauer1981quantitative, hallauer2010quantitative}. Ever since heterosis was scientifically documented by \cite{darwin1876effects}, it has been successfully manipulated to improve many species for the food, feed, and fuel industries, such as rice \citep{yu1997importance}, alfalfa \citep{riday2002heterosis}, tomatoes \citep{krieger2010flowering}, and fish \citep{wohlfarth1993heterosis}. Despite the intensive study and successful utilization of heterosis, the basic genomic mechanisms remain unclear \citep{coors1999genetics, lippman2007heterosis}. Researchers speculate that gene-expression heterosis $-$ i.e, enhanced expression of one or more genes in the hybrid relative to the inbred parents $-$ is a possible  mechanisms for the phenotypic heterosis \citep{swanson2006all, springer2007allelic}.

Recently, \cite{ji2014estimation} introduced an approach to assess gene heterosis using microarray data under the assumption that these data are continuous. They built a normal hierarchical model for gene expression that allows borrowing of information across genes for means and variances. They introduced an empirical Bayes framework that first estimates model hyperparameters, then estimates the posterior distribution for gene-specific parameters conditional on those hyperparameters, and finally computes heterosis probabilities based on integrals of regions under this posterior. This development was necessary due to the composite null hypothesis in tests for heterosis. These hypotheses, which many available methods do not fully accommodate, remain a challenge in the transition from continuous data to count data. Building on the work of \citeauthor{ji2014estimation} with the normal data model, we construct a hierarchical model based on a negative-binomial data model. We also utilize an empirical Bayes approach to obtain \emph{maximum a posterior} (MAP) estimates of the hyperparameters and the posterior distributions for the gene-specific parameters conditional on those hyperparameters.

The remainder of the paper proceeds as follows. Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of gene heterosis. Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods. Section \ref{s:maize} analyzes a maize experiment where phenotypic heterosis is well established and identifies genes demonstrating heterosis. The data and code necessary to perform this analysis is available in the Supplemental Material. Section \ref{s:discussion} summarizes the work and suggests directions for future research.

\section{Empirical Bayes identification of gene heterosis from RNAseq counts}
\label{s:method}

An RNA sequencing (RNAseq) experiment to identify genes demonstrating heterosis consists of three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H). For each variety, replicate RNA samples are isolated and assessed for quality. cDNA libraries, or collections of short cDNA fragments, are constructed. Then, next generation sequencing technology is used to determine the \emph{reads}, or nucleotide sequences, in the cDNA libraries. These reads are processed using bioinformatic algorithms to match the reads to genes, or specific gene transcripts, exons, microRNAs, etc. The result of matching reads to features are summarized by a gene $\times$ sample matrix of counts. See \cite{nettleton2014design} for more details on RNAseq experiments, and see \cite{paschold2012complementation} for the biological background behind the above steps.

To use RNAseq counts to identify genes displaying heterosis, we build a hierarchical model to borrow information on gene-variety means and gene-specific overdispersion parameters, estimate the parameters using a empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.

\subsection{Hierarchical model for RNAseq counts}
\label{s:model}

Let $Y_{gvi}$ be the count for gene $g=1,\ldots,G$, variety $v=1,\ldots,V$, and replicate $i=1,\ldots,n_v$. As shown in equation \eqref{e:data}, we assume a negative binomial for the counts with a mean that depends on the gene-variety combination through $\mu_{gv}$ and the sequencing depth $\gamma_{vi}$ and a gene-specific overdispersion $\psi_g$. 
\begin{equation} 
Y_{gvi} \stackrel{ind}{\sim} NB(e^{\mu_{gv}+\gamma_{vi}},\psi_g) 
\label{e:data}
\end{equation}
where $NB(\eta,\psi)$ indicates a negative-binomial with expectation $\eta$ and variance $\eta+\psi\eta^2$, and $ind$ indicates the observations are conditionally independent.

We present a hierarchical model for the sequencing depth, overdispersion, and gene-variety parameters. The sample-specific sequencing depth terms are assumed to follow a normal distribution, i.e. $\gamma_{vi} \stackrel{ind}{\sim} N(0,\tau^2)$. (These terms are the normalization factors that practitioners typically estimate for each sample and then treat as constant in further analysis.) The logarithm of the gene-specific overdispersion parameters and the gene-specific variety effects are assumed to have a joint normal distribution, i.e. 
$\theta_g = [\mu_g, \log(\psi_g)]^\top \stackrel{ind}{\sim} N\left(\eta, \Sigma\right)$
where $\mu_g = (\mu_{g1},\ldots,\mu_{gV})'$. We consider an \emph{independent} model in which $\Sigma$ is diagonal with elements $\sigma_1^2,\ldots,\sigma_{V+1}^2$ and a \emph{covariance} model in which we estimate all parameters in $\Sigma$.

We assume improper uniform priors for $\eta$ and take 
%In fact, the prior on \tau is just Unif(0,5)
$\tau\stackrel{ind}{\sim} Ca^+(0,3)$, where 0 and 3 are the location and scale parameters, respectively, of the untruncated half-Cauchy distribution. In the independent model, we also assign the $Ca^+(0,3)$ prior to the
diagonal elements
%square root of diagonal elements?
of $\Sigma$. In the covariance model, we assume a conjugate, proper inverse Wishart distribution with degrees of freedom $V+2$ and an identity scale matrix, resulting in marginally uniform correlations.

@@

We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ and hyperparameters $\pi = (\gamma,\eta, \Sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$. We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional of this estimated hyperparameter. Conditional on the hyperparameters, the gene-specific parameters are independent as in equation \eqref{e:condind}. Thus, conditional posterior inference for the gene-specific parameters can be obtained independently and in parallel.
\begin{align}
p\left(\theta\left|y,\hat{\pi}\right.\right) 
%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},\psi_g\right.\right) \right] N\left(\left.\left[\mu_g, \log(\psi_g) \right]^\top\right|\hat{\eta}, \hat{\Sigma} \right) 
\label{e:condind}
\end{align}

To perform the optimization, we use the statistical software, {\tt Stan} \citep{stan-software:2014}, executed through the RStan interface \citep{rstan-software:2014} in R \citep{R2014}. The Stan software allows the construction of general Bayesian models with the primary restriction that all unknown parameters are continuous. With a fully constructed model, the software can obtain MAP estimates \cite[Section 50.3,][]{stan-manual:2014} using the {\tt optimizing} function in RStan. It was necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} and reasonable starting values obtained via {\tt edgeR}, presumably due to the large dimension of the optimization. The {\tt Stan} software also includes the RStan function, {\tt sampling}, an implementation of a Markov chain Monte Carlo (MCMC) algorithm for obtaining samples from the posterior. By default, Stan uses a no-U-turn sampler \citep{hoffman2013no}, a variant of Hamiltonion Monte Carlo \citep{neal2011mcmc}.

@@

\subsection{Gene heterosis}
\label{s:gene_heterosis}

As defined in Section \ref{s:intro}, heterosis is increased (or decreased) phenotypic expression in a hybrid relative to its inbred parents. Thus, when attempting to identify heterosis, there are $V=3$ varieties of interest. We let $v=1,2$ be the two parental varieties and $v=3$ be the hybrid. We define gene heterosis to be the differential expression of a gene in the hybrid relative to its parents. For a specific gene $g$, low-parent gene heteorisis (LPH) occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$ 
%, mid-parent gene heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), 
and high-parent gene heterosis (HPH) occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{extreme gene heterosis} (EH) to be the event of LPH or HPH. Equation \eqref{e:hypotheses} shows the most important hypotheses, where $H_0$ indicates no EH and $H_1$ indicates EH. 
\begin{align}
H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}.
\label{e:hypotheses}
\end{align}
We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g. 
\begin{align}
P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
&\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right) \label{e:probs}
\end{align}
where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. The set of probabilities $P\left(H_{g1}|y, \hat{\pi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a list of genes sorted by the relative strength of evidence of extreme heterosis. Geneticists can use this list to prioritize future experiments to understand the genetic mechanism for heterosis.

\section{Simulation study based on maize experiment}
\label{s:simulation}

@@

To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation}. Section \ref{s:maize} describes this dataset in detail. We compared our method to approaches using the R packages {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes}.

@@

Initially, the maize experiment had RNAseq counts for 39,656 genes and 4 replicates per variety. We eliminated low count genes by removing any gene with three or more zeros for any variety and genes whose mean count across the varieties was less than one, resulting in 27,888 remaining genes.

Next, we applied the {\tt edgeR} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} to the maize data to obtain sample-specific normalization factors, then gene-specific dispersions, and finally gene-variety-specific location parameters. We treated these parameters as the truth for the simulation study: i.e. a gene with $\mu_{g3} < \min(\mu_{g1}, \mu_{g2})$ was considered to have LPH, while a gene with $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$ was considered to have HPH. This technique ensured that our simulated datasets maintained the existing covariance structure among the gene-variety means of the original maize data.

Using these parameters and normalization factors, we simulated data according to the negative-binomial model in equation \eqref{e:data} independently for each gene. For each simulation, low count genes $-$ those with three or more zeros in any variety or whose mean count across the varieties was less than one $-$ were removed. We retained a random subset of 25,000 genes from those with high enough counts. We repeated this simulation process 10 times for each of 4, 8, and 16 replicates per variety, reusing normalization factors when necessary.

@@

Existing RNAseq data analysis methods are not specifically designed to identify gene heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes} to provide a measure of the strength of extreme heterosis for each gene such that large values of this measure indicate support for extreme heterosis.

The {\tt edgeR} approach is designed to apply a negative-binomial log-linear model to test for differential expression between any two varieties \citep{robinson2007moderated, robinson2010edgeR}. To construct a measure of heterosis, we computed the maximum likelihood estimates of the mean expression effects of all genes using {\tt edgeR}'s built-in Fisher scoring algorithm, and then used the likelihood ratio test functionality to calculate two legitimate p-values for each gene: one for the null hypothesis that the hybrid's mean expression is the same as that of parent 1 and the second for the analogous test involving parent 2. For each gene, if the estimated mean expression level of the hybrid is between the estimated mean expression levels of the parents, we set the measure of heterosis to zero. Otherwise, we set the measure of heterosis to be one minus the p-value corresponding to the parent with estimated expression level closer to that of the hybrid.

The {\tt baySeq} approach \citep{hardcastle2010bayseq, hardcastle2012baySeq} allows for a wider range of hypotheses for each gene, including $H_1: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H_2: \mu_{g1}=\mu_{g2}$, $H_3: \mu_{g1}=\mu_{g3}$, $H_4: \mu_{g2}=\mu_{g3}$, and $H_5:$ all $\mu_g$'s are distinct. In a technique similar to our application of {\tt edgeR}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene. We set this measure to zero if the hybrid's maximum likelihood estimate of the mean expression level calculated using {\tt edgeR} is between those of the parents. Otherwise, the measure is the sum of the posterior probabilities of $H_2$ and $H_5$, the only hypotheses that allow for extreme heterosis.

>>>>

\newcommand{\blind}{0} % 0 = not blind, 1 = blind

\title{Empirical Bayes analysis of RNA-seq data for detection of gene expression heterosis}

@@

\begin{abstract}
Heterosis, or hybrid vigor, refers to the enhancements in the phenotype of hybrid progeny relative to their inbred parents. Although the heterosis phenomenon is extensively utilized in agriculture, the molecular basis of heterosis is still largely unknown. In an effort to understand phenotypic heterosis at the molecular level, researchers have begun to measure transcript abundance levels of thousands of genes in parental inbred lines and their hybrid offspring using RNA sequencing (RNA-seq) technology.  The resulting data allow researchers to search for evidence of gene expression heterosis as one potential molecular mechanism underlying heterosis of agriculturally important traits.  The null hypotheses of greatest interest in testing for gene expression heterosis are composite null hypotheses that are difficult to test with standard statistical approaches for RNA-seq analysis. To address these shortcomings, we develop a hierarchical model based on a negativeb binomial data model and draw inferences using a computationally tractable empirical Bayes approach to inference. We demonstrate improvements over alternative methods via a simulation study based on a maize experiment and then analyze that maize experiment with our newly proposed methodology. This article has supplementary material online.
\end{abstract}

\begin{keywords}
Empirical Bayes; Gene expression; Heterosis; Hierarchical model; Negative binomial; RNA-seq.
\end{keywords}

@@

Heterosis, or hybrid vigor, occurs when hybrid progeny display phenotypes that are superior to the phenotypes of their parents.  The heterosis phenomenon was scientifically documented in plants by \cite{darwin1876effects}and has long been used to improve agricultural production.  One classic example involves hybrid maize offspring that are taller, faster to mature, and yield considerably more grain than their inbred parents \citep{hallauer1981quantitative, hallauer2010quantitative}.  Despite intensive study and successful use of heterosis in agriculture, the basic molecular genetic mechanisms remain largely unknown \citep{coors1999genetics, lippman2007heterosis}. One potential explanation, known as gene expression heterosis, is enhanced expression of one or more genes in hybrids compared to their inbred parents \citep{swanson2006all, springer2007allelic}.

Recently, \cite{ji2014estimation} introduced an approach to assess gene expression heterosis using microarray data under the assumption that these data are continuous. They built a normal hierarchical model for microarray measurements of transcript abundance that allows borrowing of information across genes to estimate means and variances. They introduced an empirical Bayes framework that first estimates model hyperparameters, then estimates the posterior distribution for gene-specific parameters conditional on those hyperparameters, and finally computes heterosis probabilities based on integrals of regions under this posterior. This development was necessary due to the composite null hypotheses in tests for heterosis. These hypotheses, which many available methods do not fully accommodate, remain a challenge in the transition from continuous measurements of transcript abundance to count-based measurements that arise from RNA sequencing (RNA-seq) technology. Building on the work of \citeauthor{ji2014estimation} with the normal data model, we construct a hierarchical model based on a negative-binomial data model. We also utilize an empirical Bayes approach to obtain \emph{maximum a posterior} (MAP) estimates of the hyperparameters and the posterior distributions for the gene-specific parameters conditional on those hyperparameters.

The remainder of the paper proceeds as follows. Section \ref{s:method} presents the proposed hierarchical model, an empirical Bayes method of estimating the parameters, and the calculation of posterior probabilities of gene expression heterosis. Section \ref{s:simulation} presents a simulation study based on a maize experiment and compares our approach to alternative methods. Section \ref{s:maize} analyzes a maize experiment where phenotypic heterosis is well established and identifies genes demonstrating expression heterosis. The data and code necessary to perform this analysis are available in the Supplemental Material. Section \ref{s:discussion} summarizes the work and suggests directions for future research.

\section{Empirical Bayes identification of gene expression heterosis from RNA-seq read counts}
\label{s:method}

An RNA sequencing (RNA-seq) experiment to identify genes demonstrating expression heterosis consists of three genetic varieties: two parental varieties (P1 and P2) and a cross between these two varieties called the hybrid (H). For each variety, replicate RNA samples are isolated and assessed for quality. Complementary DNA (cDNA) libraries, consisting of short cDNA fragments derived from RNA, are constructed. Then, next generation sequencing technology is used to determine the \emph{reads}, or nucleotide sequences, in the cDNA libraries. These reads are processed using bioinformatic algorithms to match the reads to genes, or specific gene transcripts, exons, microRNAs, etc. The results of read processing are summarized by a gene $\times$ sample matrix of counts. See \cite{nettleton2014design} 
% JBN: cite the book 
for more details on RNA-seq experiments and data from a statistical perspective, and see \cite{paschold2012complementation} for the biological background behind the use of RNA-seq to study gene expression heterosis.

To use RNA-seq counts to identify genes displaying expression heterosis, we build a hierarchical model to borrow information across gene-variety means and across gene-specific overdispersion parameters, estimate the parameters using an empirical Bayes procedure implemented in the statistical software {\tt Stan}, and calculate empirical Bayes posterior probabilities for heterosis.

\subsection{Hierarchical model for RNA-seq counts}
\label{s:model}

Let $Y_{gvi}$ be the count for gene $g=1,\ldots,G$, variety $v=1,\ldots,V$, and replicate $i=1,\ldots,n_v$.  We assume
\begin{equation} 
Y_{gvi} \stackrel{ind}{\sim} NB(e^{\mu_{gv}+\gamma_{vi}},\psi_g), 
\label{e:data}
\end{equation}
where $NB(\eta,\psi)$ indicates a negative binomial distribution with expectation $\eta$ and variance $\eta+\psi\eta^2$, and $ind$ indicates the observations are conditionally independent.  As shown in equation \eqref{e:data}, our data model involves gene-specific overdispersion $\psi_g$ and a mean that depends on the gene-variety combination through $\mu_{gv}$ and on the sample though  $\gamma_{vi}$.  The $\mu_{gv}$ terms are of primary scientific interest; the $\gamma_{vi}$ terms are normalization factors that account for differences in the thoroughness of sequencing from sample to sample.  Practitioners typically compute normalization factors for each sample and then treat them as known constants in further analysis.  In contrast, we include normalization factors in our model as unknowns that we estimate along with other parameters.

We present a hierarchical model for the normalization factors, overdispersions, and gene-variety parameters. The sample-specific sequencing depth terms are assumed to follow a normal distribution, i.e. $\gamma_{vi} \stackrel{ind}{\sim} N(0,\tau^2)$. (These terms are the normalization factors that practitioners typically estimate for each sample and then treat as constant in further analysis.) The logarithm of the gene-specific overdispersion parameters and the gene-specific variety effects are assumed to have a joint normal distribution, i.e. 
$\theta_g = [\mu_g, \log(\psi_g)]^\top \stackrel{ind}{\sim} N\left(\eta, \Sigma\right)$
where $\mu_g = (\mu_{g1},\ldots,\mu_{gV})'$. We consider an \emph{independence} model in which $\Sigma$ is diagonal with elements $\sigma_1^2,\ldots,\sigma_{V+1}^2$ and a \emph{dependence} model in which $\Sigma$ is a general positive definite matrix.
% JBN: Switch mu and eta throughout

We assume improper uniform priors for $\eta$ and take 
%In fact, the prior on \tau is just Unif(0,5)
$\tau\stackrel{ind}{\sim} Ca^+(0,3)$, where 0 and 3 are the location and scale parameters, respectively, of the untruncated half-Cauchy distribution. In the independence model, we also assign the $Ca^+(0,3)$ prior to the
diagonal elements
%square root of diagonal elements?
of $\Sigma$. In the dependence model, we assume a conjugate, proper inverse Wishart distribution with degrees of freedom $V+2$ and an identity scale matrix, resulting in marginally uniform correlations.

@@

We separate the parameters of the model in Section \ref{s:model} into gene-specific parameters $\theta = (\theta_1,\ldots,\theta_G)$ and hyperparameters $\pi = (\gamma,\eta, \Sigma)$, where $\gamma = (\gamma_{11},\ldots,\gamma_{Vn_V})$. We employ an optimization procedure to obtain the \emph{maximum a posteriori} (MAP) estimate for $\pi$, denoted $\hat{\pi}$, and then base gene-specific inference on the posterior conditional on $\hat{\pi}$. Conditional on the hyperparameters, the gene-specific parameters are independent, i.e., 
\begin{align}
p\left(\theta\left|y,\hat{\pi}\right.\right) 
%&= \prod_{g=1}^G p\left(\theta_g\left|y_g,\hat{\pi}\right.\right) = \prod_{g=1}^G p\left(y_g\left|\theta_g,\hat{\gamma}\right.\right) p\left(\theta_g\left|\hat{\pi}\right.\right) \nonumber \\
&\propto \prod_{g=1}^G \left[ \prod_{v=1}^V \prod_{i=1}^{n_v} NB\left(y_{gvi}\left|e^{\mu_{gv}+\hat{\gamma}_{vi}},\psi_g\right.\right) \right] N\left(\left.\left[\mu_g, \log(\psi_g) \right]^\top\right|\hat{\eta}, \hat{\Sigma} \right). 
\label{e:condind}
\end{align}
Thus, conditional posterior inference for the gene-specific parameters can be obtained independently and in parallel.

To perform the optimization required for the MAP estimate $\hat{\pi}$, we use the statistical software, {\tt Stan} \citep{stan-software:2014}, executed through the RStan interface \citep{rstan-software:2014} in R \citep{R2014}. The Stan software allows the construction of general Bayesian models with the primary restriction that all unknown parameters are continuous. With a fully constructed model, the software can obtain MAP estimates \cite[Section 50.3,][]{stan-manual:2014} using the {\tt optimizing} function in RStan. Presumably due to the large dimension of the optimization, it was necessary to use the L-BFGS algorithm \cite[Section 55,][]{stan-manual:2014} and reasonable starting values obtained via {\tt edgeR} \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor}. The {\tt Stan} software also includes the RStan function, {\tt sampling}, an implementation of a Markov chain Monte Carlo (MCMC) algorithm for obtaining samples from the posterior. By default, Stan uses a no-U-turn sampler \citep{hoffman2013no}, a variant of Hamiltonion Monte Carlo \citep{neal2011mcmc}.

@@

\subsection{Gene expression heterosis}
\label{s:gene_heterosis}

To identify gene expression heterosis, we consider $V=3$ varieties, where  $v=1,2$ signify the two parental varieties and $v=3$ represents the hybrid. For a specific gene $g$, low-parent  heteorisis (LPH) of gene expression occurs when $\mu_{g3}< \min(\mu_{g1},\mu_{g2})$ 
%, mid-parent gene expression heterosis occurs when $\mu_{g1}\le \mu_{g3}\le \mu_{g2}$ if $\mu_{g1}\le \mu_{g2}$ (or $\mu_{g1}\ge \mu_{g3}\ge \mu_{g2}$ if $\mu_{g1}\ge \mu_{g2}$), 
and high-parent  heterosis (HPH) of gene expression occurs when $\max(\mu_{g1},\mu_{g2}) < \mu_{g3}$. For simplicity, we define \emph{ heterosis} of gene expression to be the event of LPH or HPH. The most important hypotheses are
\begin{align}
H_{g0}:&\mu_{g1}\le \mu_{g3}\le \mu_{g2} \mbox{ or } \mu_{g1}\ge \mu_{g3}\ge \mu_{g2} \mbox{\ \ vs.\ \ } \nonumber \\
H_{g1}:&\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3},
\label{e:hypotheses}
\end{align}
where $H_{g0}$ indicates no heterosis for gene $g$ and $H_{g1}$ indicates  heterosis for gene $g$. 
We evaluate these hypotheses based on empirical Bayes estimates of their posterior probabilities, e.g., 
\begin{align}
P\left(H_{g1}|y, \hat{\pi}_{MAP}\right) &= P\left(\left.\mu_{g3}< \min(\mu_{g1},\mu_{g2}) \mbox{ or } \max(\mu_{g1},\mu_{g2}) < \mu_{g3}\right| y, \hat{\pi}_{MAP}\right) \nonumber \\
&\approx \frac{1}{M} \sum_{m=1}^M \mathrm{I}\left(\mu_{g3}^{(m)}< \min\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) \mbox{ or } \max\left(\mu_{g1}^{(m)},\mu_{g2}^{(m)}\right) < \mu_{g3}^{(m)}\right), \label{e:probs}
\end{align}
where $\mu_g^{(m)} = \left(\mu_{g1}^{(m)},\mu_{g2}^{(m)},\mu_{g3}^{(m)}\right)$ is the $m^{th}$ MCMC sample from the empirical Bayes posterior, and $\mathrm{I}(A)$ is 1 if A is true and 0 otherwise. The set of probabilities $P\left(H_{g1}|y, \hat{\pi}_{MAP}\right)$ for $g=1,\ldots,G$ can be used to provide a list of genes sorted by the relative strength of evidence of  heterosis. Geneticists can use this list to prioritize future experiments to understand the molecular genetic mechanisms for heterosis.

\section{Simulation study based on a maize experiment}
\label{s:simulation}

@@

To assess the efficacy of our method to identify genes demonstrating heterosis, we used a maize data set with parental varieties B73 and Mo17 and the hybrid variety (B73 $\times$ Mo17) \citep{paschold2012complementation} to determine realistic parameter values for a simulation study. Section \ref{s:maize} describes the maize dataset in detail. We compared our method to approaches using the R packages {\tt edgeR}, {\tt baySeq} \citep{hardcastle2010bayseq, hardcastle2012baySeq}, and {\tt ShrinkBayes} \citep{van2014shrinkbayes}.

@@

We applied the {\tt edgeR} package \citep{robinson2010edgeR} from Bioconductor \citep{gentleman2004bioconductor} to the maize data on 27,888 genes with average count at least one and at most two zero read counts for each variety across four replications.  This analysis produced  sample-specific normalization factors () % JBN: state the normalization factors, "I suppose you used the edgeR default, which is TMM, but I think this should be stated explicitly."
, gene-specific dispersion estimates, and gene-variety-specific location parameter estimates. We treated these estimates as the true parameter values for the simulation study so that our simulated datasets mimicked the existing covariance structure among the gene-variety means of the original maize data.

Using these parameters and normalization factors, we simulated data according to the negative-binomial model in \eqref{e:data} independently for each gene. For each simulation, we analyzed a subset of 25,000 genes selected randomly from genes with simulated counts at least one on average and with at most two zeros for each variety across replications.  We repeated this simulation process 10 times for each of 4, 8, and 16 replicates per variety, reusing normalization factors when necessary.  
% From Dan "Did we consider changing the number of zeros allowed when the number of reps is higher?  I like the idea of simplifying all this by just picking an average count threshold that his high enough to avoid problems.  I think that may be what others have done in the past."

@@

Existing RNA-seq data analysis methods are not specifically designed to identify gene expression heterosis, but we adapted the approaches in {\tt edgeR}, {\tt baySeq}, and {\tt ShrinkBayes} to provide a measure of the strength of heterosis for each gene such that large values of this measure indicate support for heterosis.

The {\tt edgeR} approach can be used to test for differential expression between any two varieties based on the fit of a negative binomial log-linear model \citep{robinson2007moderated, robinson2010edgeR}. To construct a measure of heterosis, we computed the maximum likelihood estimates of the $\mu_{gv}$ parameters for all genes using {\tt edgeR}'s built-in Fisher scoring algorithm, and then used likelihood ratio tests to calculate two $p$-values for each gene: $p_{g1}$  for testing $H_{g01}:\mu_{g1}=\mu_{g3}$ and $p_{g2}$ for testing $H_{g02}:\mu_{g2}=\mu_{g3}$. Then, for each gene, we define a new $p$-value denoted as $p_g$ and determined as $p_g=1$ if the estimate of $\mu_{g3}$ falls between the estimates of $\mu_{g1}$ and $\mu_{g2}$ and $p_g=\max\{p_{g1},p_{g2}\}/2$ otherwise.  (For all relevant signifcance thresholds $\omega$ near 0, it can be shown that rejecting the null hypothesis of no heterosis of expression for gene $g$ whenever $p_g\leq \omega$ results in a test that is asymptotically size $\omega$.)  We then use $1-p_g$ as a final measure positively associated with strength of evidence for gene expression heterosis.

The {\tt baySeq} approach allows for a wider range of hypotheses for each gene, including $H^*_{g1}: \mu_{g1}=\mu_{g2}=\mu_{g3}$, $H^*_{g 2}: \mu_{g1}=\mu_{g2}$, $H^*_{g 3}: \mu_{g1}=\mu_{g3}$, $H^*_{g 4}: \mu_{g2}=\mu_{g3}$, and $H^*_{g 5}:$ all $\mu_g$'s are distinct. In a technique similar to our application of {\tt edgeR}, we used the posterior probabilities of these hypotheses to construct a measure of heterosis for each gene. We set this measure to zero if the maximum likelihood estimate, calculated using {\tt edgeR}, of $\mu_{g3}$ is between the maximum likelihood estimates of $\mu_{g1}$ and $\mu_{g2}$. Otherwise, the measure is the sum of the posterior probabilities of $H^*_{g 2}$ and $H^*_{g 5}$, the only hypotheses that allow for heterosis.

The {\tt ShrinkBayes} package implements empirical Bayesian inference via integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}. A limitation of INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior. Thus, we cannot directly estimate posterior probabilities involving nonlinear functions of $\mu_{g1}$, $\mu_{g2}$, and $\mu_{g3}$, such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2})|y)$. To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} using the approach in \cite{ji2014estimation}. 
Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the average of parental effects, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental effect difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect relative to the average parental effect. After fitting the model with {\tt ShrinkBayes}, we constructed a measure of heterosis for each gene. For gene $g$, we used the estimated posterior means $\hat{\delta}_g$ and $\hat{\alpha}_g$ of $\delta_g$ and $\alpha_g$, respectively, along with appropriate posterior probabilities. Specifically, the measure of heterosis for gene $g$ is zero if $|\hat{\delta}_g| \le |\hat{\alpha}_g|$, $P(\delta_g + \alpha_g > 0 |y)$ if $\hat{\delta}_g > -\hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g > 0 |y)$ if $\hat{\delta}_g > \hat{\alpha}_g \ge 0$, $P(\delta_g - \alpha_g < 0 |y)$ if $\hat{\delta}_g < \hat{\alpha}_g \le 0$, and $P(\delta_g + \alpha_g < 0 |y)$ if $\hat{\delta}_g < -\hat{\alpha}_g \le 0$.
